A weakly structured stem for human origins in Africa

### 2022-03-11 paper.tex

\begin{abstract}
While it is now broadly accepted that \emph{Homo sapiens}
origined within Africa, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in prior dating estimates.
Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models \comment{\ldots to evaluate the fit of various models}.
We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, contemporary
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135ka, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral \emph{Homo}
populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly structured stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar.
Despite genetic similarity between these populations, an inferred 1--4\%
\sgcomment{explain in the main text where this estimate is from!}
of genetic differentiation among contemporary human populations can be
attributed to genetic drift between stem populations.
We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
\end{abstract}

@@

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}. These
approaches are more challenging to interpret and fit to data.
\comment{I think a schematic illustrating models being tested would be extremely useful in the main text. This could be Fig S2, or even a new figure with a further simplified illustration o fthe general idea of the models being tested to help orient the reader.}
However, new
population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent articles
have shown that introducing an archaic hominin ghost population contributing to African
populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models
\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Durvasula2020-td,Lorente-Galdos2019-vz,Durvasula2020-td}.
This has driven speculation about the geographic range of this ghost
population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence \citep{Hsieh2016-gk}). However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic hominin admixture model, leaving out other plausible models
\citep{Henn2018-rf} (Figure~\ref{fig:supp-possible-models}).
Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African or Bantu-speaking ancestry (Figure \ref{fig:1}C). 
While ancient DNA from Eurasia has helped us understand early human history
outside of Africa, there is no comparably ancient DNA to elucidate early history
in Africa \citep{Lipson2022-xf}.

Here, we therefore aim to discriminate among a broader set of demographic
models by interrogating the genomes of contemporary populations. We take as our
starting point 4 classes of models (single population expansion, single
population expansion with regional persistence, archaic hominin admixture, and
multi-regional evolution, Figure~\ref{fig:supp-possible-models}), using 290
genomes from southern, eastern, and western Africa as well as Eurasia. By
including geographically and genetically diverse populations across Africa, we
infer demographic models that explain more features of genetic diversity
in more
populations than previously reported. These analyses confirm the inadequacy of
tree-like models and provide an opportunity to directly evaluate a wide range
of alternative models.

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (from Sierre Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendents of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2019-wd}), and Eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2019-wd} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for gene flow from Neanderthals into non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged 550ka from a common
stem. We computed one- and two-locus statistics whose expectation within and across 
populations can be computed efficiently and that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years. \sgcomment{define LD}

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/fig1.pdf}
    \caption[width=\textwidth]{
        \textbf{Genetic diversity across Africa.}
        (A) \sgcomment{secure coordinates} Select populations from the 1000 Genomes and African Diversity Reference Panels
        illustrate diversity from western, eastern and southern Africa.
        We chose representatives from each region (bold labels)
        to build parameterized models,
        including the newly-sequenced Nama from South Africa, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderthal individual.
        (B, C) PCA highlights the range of genetic divergence anchored 
        by western Africans, Nama, Gumuz and the British.
        Percentages show variance explained by each principal component.
        (D) \texttt{ADMIXTURE} with $K=4$ illustrates signatures of recent gene flow in Africa
        which reflect colonial-period migration into the Nama,
        Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
        %\comment{How did you land on the best fit k? Show scree plot of CV errors, or otherwise justify choice}
    }
    \label{fig:1}
\end{figure}

@@

We began with a model of geographic expansion from a single ancestral,
unstructured source followed by migration between populations, without allowing
for contribution from an African archaic hominin lineage or population structure prior
to the expansion (Figure~\ref{fig:supp-possible-models}A). As expected
\citep{Ragsdale2019-nt}, this first model was a poor fit to the data
qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and quantitatively
(log-likelihood ($LL$) $\approx -189,400$, Table~\ref{tab:supp-single-origin}).
We next explored a suite of models in which population structure predates the
differentiation of contemporary groups, including models allowing for ancestral
reticulation (such as fragmentation-and-coalescence or meta-population models,
Figure~\ref{fig:supp-possible-models}B), archaic hominin admixture
(Figure~\ref{fig:supp-possible-models}C), and African multi-regionalism
(Figure~\ref{fig:supp-possible-models}D).

Regardless of the model choice for early epochs, inference of human demographic
history for the last 150ka was remarkably robust. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135ka, with low to moderate levels of
subsequent gene flow (Table \ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140ka. 
We conclude that geographic patterns of contemporary \emph{Homo sapiens} population structure
date back to the late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), contemporary populations cannot be easily mapped onto the more ancient
`stem' groups as only a small proportion of drift between contemporary populations can
be attributed to drift between stems (Figures~\ref{fig:3} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events. 
Fixed parameters included
the time of divergence between Western and Eastern African populations,
set to 60ka, just prior to the split of Eurasians and East Africans set to 50ka. 
We also fixed the amount of admixture from Neanderthals to Europeans directly
following the out-of-Africa migration which was set to 1.5\% at 45ka (Supp. Information). These constraints allowed us to integrate information from previous genetic and archaelogical research,
to infer robust migration rates. For example, all models infer relatively high gene flow between Eastern and Western Africa
($m\approx2\times10^{-4},$ the proportion of migrant lineages per generation). We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE} (Figure~\ref{fig:1}).

\begin{figure}[t!]
    \centering
    \includegraphics[width=0.65\textwidth]{figures/fig2-melisa}
    \caption{
        \textbf{A weakly structured stem best describes two-locus statistics}
        In the two best-fitting parameterizations of early population structure,
        continuous migration (A) and multiple mergers (B), models
        that include ongoing migration between stem populations outperform
        those in which stem populations are isolated. Most recent populations are
        connected by continuous, reciprocal migration that are indicated by 
        double-headed arrows (labels matched to migration rates and divergence
        times in Table~\ref{tab:migration-rates}). These migrations last for the
        duration of co-existence of contemporaneous populations with constant
        migration rates over those intervals. The
        merger-with-stem-migration model (B, with  $LL=-102,600$) outperformed the
        continuous-migration model (A, with $LL=-115,500$ ).
        Colors are used to distinguish overlapping branches and link to Figure \ref{fig:1}.}
    \label{fig:2}
\end{figure}

\begin{table}[t!]
    \centering
    \scriptsize
    \begin{tabular}{lllrrrr}
        \toprule
        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
        & & & 
            \textbf{Time (ka)} & \textbf{per generation} & \textbf{duration (kyr)} \\
        \midrule
        \textbf{Continuous Model} & & & & & \\
        $LL= -115,500$ & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
        & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
        & g & Mende, East Africans & 60 & \textbf{2.14e-4} & 60 \\
        & h & East Africans, British & 50 & 4.17e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.36e-4} & 12 \\
        \textbf{Merger Model} & & & & & \\
        $LL= -102,600$ & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
        & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
        & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
        & c & Stem 2 to Stem 1E & 98 & \textbf{0.50} & pulse \\
        & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
        & e, f & Nama, Other Africans* & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
        & g & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
        & h & East Africans, British & 50 & 3.87e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
        \bottomrule
    \end{tabular}
    \caption{
        \textbf{Migration and divergence parameters from best fit models.}
        Labeled migration rates correspond to symmetric continuous migration
        bands shown in Figure~\ref{fig:2}. Both the continuous migration and
        merger models inferred a relatively deep split of human stem branches,
        though they were connected by ongoing migration that maintained their
        genetic similarity. Bold text indicates migration rates above $10^{-4}$.
        In both models, the branch ancestral to the Nama shares a common ancestral population
        with the other African groups $\sim$120--135ka. Following this divergence,
        the population ancestral to other African groups branches into West and East African
        groups 60ka.
        *Migration rates and durations are shown between branches ancestral to 
        1) Nama and East Africans and their ancestors, and
        2) Nama and Mende, respectively.
        ``Divergence times'' correspond to the most recent common ancestral population,
        and does not account for continuous migration or earlier reticulations.
    }
    \label{tab:migration-rates}
\end{table}

@@

To account for population structure prior to 135ka, three of our four models allowed for two or more
``stem'' populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase. 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or ``merger'') events from the
other stem population, so that recent populations are formed by mergers of multiple ancestral populations. Depending on parameter values,
this scenario encompasses archaic hominin introgression and
fragmentation-and-coalescence models.
\comment{another supp schematic figure may be helpful}
For many parameters, confidence intervals based on bootstrapping are relatively narrow
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. Model assumptions
have, however, a larger impact on parameter estimates (and thus, real uncertainty). To convey model uncertainty, we
highlight features of the two inferred models with high likelihoods. These are referred to as the
``multiple-merger'' and the ``continuous-migration'' models. Both allow for
migration between stem branches, but differ primarily in the timing of the
early divergence of stem populations and their relative $N_e$
(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
during the Middle Stone Age.

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration ($LL \approx -102,600$ vs.
$-107,700$ in the merger model and $LL \approx -115,500$ vs. $-126,600$ with
continuous migration). With continuous migration between stems, population
structure extends back to 1.1--1.4Ma (Table~\ref{tab:migration-rates}).
Migration between the stems in these models is moderate, with a fraction of
migrant lineages each generation estimated as
$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is similar
to inferred migration rates between connected contemporary populations over the
past 50ka (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic hominin admixture models (Figure~\ref{fig:supp-possible-models}C) as
the early branches remain closely related, and each branch contributes large amounts to all contemporary populations (Figure 3). 
Because of this 
relatedness, only $1\%$ to $4\%$ of genetic differentiation among contemporary populations can be traced back
to this early population structure (Figure 3 and Supp. Information)

@@

\begin{figure}[t!]
    \centering
    \includegraphics{figures/fig3.pdf}
    \caption{
        \textbf{Structure among stems is weak and present-day structure is mostly recent}
        From the best fit models of our two parameterizations (A and B:
        continuous migration, C and D: merger with stem migration),
        we predicted pairwise differences $H_{i,j}$ between individuals
        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
        (B and D) To understand how drift between stems explains contemporary structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of
        sampled contemporary
        populations that aligns with drift between past populations
        (here Nama and Mende, see Section~S\ref{sec:f4} for details and additional
        comparisons in Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        contemporary genetic differentiation.
        Most present differentiation dates back to the last 100ka.
    }
    \label{fig:3}
\end{figure}

@@

We are grateful for the DNA contribution from each participant which enabled
this study; in particular we wish to highlight the generous participation of
the Richtersveld Nama community in South Africa and help from local research
assistants Willem DeKlerk and Hendrik Kaimann. Additional assistance and
community engagement was conducted by Justin Myrick, Chris Gignoux, Caitlen
Uren and Cedric Werely. We thank the African Genome Diversity Project for data
generation, including Tommy Carensten, Deepti Gurdasani, and Manj Sandhu. We
thank Luke Anderson-Trocm\'e for assistance in creating the map in
Figure~\ref{fig:1}. This research was supported by an NIH grant R35GM133531 (to
BMH). The content is solely the responsibility of the authors and does
not necessarily represent the official views of the National Institutes of
Health.

>>>>

\begin{abstract}
While it is now broadly accepted that \emph{Homo sapiens}
origined within Africa, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in prior dating estimates.
Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models.
We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, contemporary
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135ka, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral \emph{Homo}
populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly structured stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar.
Despite genetic similarity between these populations, an inferred 1--4\%
of genetic differentiation among contemporary human populations can be
attributed to genetic drift between stem populations.
We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
\end{abstract}

@@

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}. These
approaches are more challenging to interpret and fit to data.
However, new population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent articles
have shown that introducing an archaic hominin ghost population contributing to African
populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models
\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Durvasula2020-td,Lorente-Galdos2019-vz,Durvasula2020-td}.
This has driven speculation about the geographic range of this ghost
population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence \citep{Hsieh2016-gk}). However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic hominin admixture model, leaving out other plausible models
\citep{Henn2018-rf} (Figure~\ref{fig:proposed-models}).
Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African or Bantu-speaking ancestry
(Figure~\ref{fig:diversity}C). 
While ancient DNA from Eurasia has helped us understand early human history
outside of Africa, there is no comparably ancient DNA to elucidate early history
in Africa \citep{Lipson2022-xf}.

Here, we therefore aim to discriminate among a broader set of demographic
models by interrogating the genomes of contemporary populations. We take as our
starting point 4 classes of models (single population expansion, single
population expansion with regional persistence, archaic hominin admixture, and
multi-regional evolution, Figure~\ref{fig:proposed-models}), using 290
genomes from southern, eastern, and western Africa as well as Eurasia. By
including geographically and genetically diverse populations across Africa, we
infer demographic models that explain more features of genetic diversity
in more
populations than previously reported. These analyses confirm the inadequacy of
tree-like models and provide an opportunity to directly evaluate a wide range
of alternative models.

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.5\textwidth]{figures/proposed-models.png}
    \caption{
        \textbf{Proposed models of early human history in Africa.}
        seeking to reconcile genetic and paleoanthropological evidence \citep{Henn2018-rf}.
        \aprcomment{needs a caption}
    }
    \label{fig:proposed-models}
\end{figure}

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (from Sierre Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendents of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2022-pw}), and Eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2022-pw} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:diversity}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for gene flow from Neanderthals into non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged 550ka from a common
stem. We computed one- and two-locus statistics whose expectation within and across 
populations can be computed efficiently and that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/samples-diversity.pdf}
    \caption[width=\textwidth]{
        \textbf{Genetic diversity across Africa.}
        (A) Select populations from the 1000 Genomes and African Diversity Reference Panels
        illustrate diversity from western, eastern and southern Africa.
        We chose representatives from each region (bold labels)
        to build parameterized models,
        including the newly-sequenced Nama from South Africa, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderthal individual.
        (B, C) PCA highlights the range of genetic divergence anchored 
        by western Africans, Nama, Gumuz and the British.
        Percentages show variance explained by each principal component.
        (D) \texttt{ADMIXTURE} with $K=4$ illustrates signatures of recent gene flow in Africa
        which reflect colonial-period migration into the Nama,
        Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
    }
    \label{fig:diversity}
\end{figure}

@@

We began with a model of geographic expansion from a single ancestral,
unstructured source followed by migration between populations, without allowing
for contribution from an African archaic hominin lineage or population structure prior
to the expansion (Figure~\ref{fig:proposed-models}A). As expected
\citep{Ragsdale2019-nt}, this first model was a poor fit to the data
qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and quantitatively
(log-likelihood ($LL$) $\approx -189,400$, Table~\ref{tab:supp-single-origin}).
We next explored a suite of models in which population structure predates the
differentiation of contemporary groups, including models allowing for ancestral
reticulation (such as fragmentation-and-coalescence or meta-population models,
Figure~\ref{fig:proposed-models}B), archaic hominin admixture
(Figure~\ref{fig:proposed-models}C), and African multi-regionalism
(Figure~\ref{fig:proposed-models}D).

Regardless of the model choice for early epochs, inference of human demographic
history for the last 150ka was remarkably robust. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135ka, with low to moderate levels of
subsequent gene flow (Table \ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140ka. 
We conclude that geographic patterns of contemporary \emph{Homo sapiens} population structure
date back to the late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), contemporary populations cannot be easily mapped onto the more ancient
`stem' groups as only a small proportion of drift between contemporary populations can
be attributed to drift between stems (Figures~\ref{fig:predictions} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events. 
Fixed parameters included
the time of divergence between Western and Eastern African populations,
set to 60ka, just prior to the split of Eurasians and East Africans set to 50ka. 
We also fixed the amount of admixture from Neanderthals to Europeans directly
following the out-of-Africa migration which was set to 1.5\% at 45ka (Supp. Information). These constraints allowed us to integrate information from previous genetic and archaelogical research,
to infer robust migration rates. For example, all models infer relatively high gene flow between Eastern and Western Africa
($m\approx2\times10^{-4},$ the proportion of migrant lineages per generation). We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE} (Figure~\ref{fig:diversity}).

\begin{figure}[t!]
    \centering
    \includegraphics[width=0.65\textwidth]{figures/best-fit-models}
    \caption{
        \textbf{A weakly structured stem best describes two-locus statistics.}
        In the two best-fitting parameterizations of early population structure,
        continuous migration (A) and multiple mergers (B), models
        that include ongoing migration between stem populations outperform
        those in which stem populations are isolated. Most recent populations are
        connected by continuous, reciprocal migration that are indicated by 
        double-headed arrows (labels matched to migration rates and divergence
        times in Table~\ref{tab:migration-rates}). These migrations last for the
        duration of co-existence of contemporaneous populations with constant
        migration rates over those intervals. The
        merger-with-stem-migration model (B, with  $LL=-102,600$) outperformed the
        continuous-migration model (A, with $LL=-115,500$ ).
        Colors are used to distinguish overlapping branches and link to
        Figure~\ref{fig:diversity}.}
    \label{fig:best-fit-models}
\end{figure}

\begin{table}[t!]
    \centering
    \scriptsize
    \begin{tabular}{lllrrrr}
        \toprule
        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
        & & & 
            \textbf{Time (ka)} & \textbf{per generation} & \textbf{duration (kyr)} \\
        \midrule
        \textbf{Continuous Model} & & & & & \\
        $LL= -115,500$ & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
        & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
        & g & Mende, East Africans & 60 & \textbf{2.14e-4} & 60 \\
        & h & East Africans, British & 50 & 4.17e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.36e-4} & 12 \\
        \textbf{Merger Model} & & & & & \\
        $LL= -102,600$ & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
        & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
        & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
        & c & Stem 2 to Stem 1E & 98 & \textbf{0.50} & pulse \\
        & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
        & e, f & Nama, Other Africans* & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
        & g & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
        & h & East Africans, British & 50 & 3.87e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
        \bottomrule
    \end{tabular}
    \caption{
        \textbf{Migration and divergence parameters from best fit models.}
        Labeled migration rates correspond to symmetric continuous migration
        bands shown in Figure~\ref{fig:best-fit-models}. Both the continuous migration and
        merger models inferred a relatively deep split of human stem branches,
        though they were connected by ongoing migration that maintained their
        genetic similarity. Bold text indicates migration rates above $10^{-4}$.
        In both models, the branch ancestral to the Nama shares a common ancestral population
        with the other African groups $\sim$120--135ka. Following this divergence,
        the population ancestral to other African groups branches into West and East African
        groups 60ka.
        *Migration rates and durations are shown between branches ancestral to 
        1) Nama and East Africans and their ancestors, and
        2) Nama and Mende, respectively.
        ``Divergence times'' correspond to the most recent common ancestral population,
        and does not account for continuous migration or earlier reticulations.
    }
    \label{tab:migration-rates}
\end{table}

@@

To account for population structure prior to 135ka, three of our four models allowed for two or more
``stem'' populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase. 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or ``merger'') events from the
other stem population, so that recent populations are formed by mergers of multiple ancestral populations. Depending on parameter values,
this scenario encompasses archaic hominin introgression and
fragmentation-and-coalescence models (such as Figure~\ref{fig:proposed-models}A and C).
For many parameters, confidence intervals based on bootstrapping are relatively narrow
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. Model assumptions
have, however, a larger impact on parameter estimates (and thus, real uncertainty). To convey model uncertainty, we
highlight features of the two inferred models with high likelihoods. These are referred to as the
``multiple-merger'' and the ``continuous-migration'' models. Both allow for
migration between stem branches, but differ primarily in the timing of the
early divergence of stem populations and their relative $N_e$
(Figure~\ref{fig:best-fit-models}). The two models also differ in the mode of divergence
during the Middle Stone Age.

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration ($LL \approx -102,600$ vs.
$-107,700$ in the merger model and $LL \approx -115,500$ vs. $-126,600$ with
continuous migration). With continuous migration between stems, population
structure extends back to 1.1--1.4Ma (Table~\ref{tab:migration-rates}).
Migration between the stems in these models is moderate, with a fraction of
migrant lineages each generation estimated as
$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is similar
to inferred migration rates between connected contemporary populations over the
past 50ka (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic hominin admixture models (Figure~\ref{fig:proposed-models}C) as
the early branches remain closely related, and each branch contributes large
amounts to all contemporary populations (Figure~\ref{fig:predictions}). 
Because of this relatedness, only $1\%$ to $4\%$ of genetic differentiation
among contemporary populations can be traced back
to this early population structure (Supp. Information)

@@

\begin{figure}[t!]
    \centering
    \includegraphics{figures/predictions.pdf}
    \caption{
        \textbf{Structure among stems is weak and present-day structure is mostly recent.}
        From the best fit models of our two parameterizations (A and B:
        continuous migration, C and D: merger with stem migration),
        we predicted pairwise differences $H_{i,j}$ between individuals
        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
        (B and D) To understand how drift between stems explains contemporary structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of
        sampled contemporary
        populations that aligns with drift between past populations
        (here Nama and Mende, see Section~S\ref{sec:f4} for details and additional
        comparisons in Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        contemporary genetic differentiation.
        Most present differentiation dates back to the last 100ka.
    }
    \label{fig:predictions}
\end{figure}

@@

We are grateful for the DNA contribution from each participant which enabled
this study; in particular we wish to highlight the generous participation of
the Richtersveld Nama community in South Africa and help from local research
assistants Willem DeKlerk and Hendrik Kaimann. Additional assistance and
community engagement was conducted by Justin Myrick, Chris Gignoux, Caitlen
Uren and Cedric Werely. We thank the African Genome Diversity Project for data
generation, including Tommy Carensten, Deepti Gurdasani, and Manj Sandhu. We
thank Luke Anderson-Trocm\'e for assistance in creating the map in
Figure~\ref{fig:diversity}. This research was supported by an NIH grant R35GM133531 (to
BMH). The content is solely the responsibility of the authors and does
not necessarily represent the official views of the National Institutes of
Health.
